Parameters estimation via HOSVD

### 2024-10-16 Code/R/Research/HOSVD parameters estimate.Rmd
Added comparisons with noise and nd case

---
title: "Parameters estimation via HOSVD"
output: html_document
---
```{r setup, include=FALSE}
library(Rssa)
library(rTensor)
```

```{r tensorization}
tens3 <- function(s, I, L, kind = c("HO-SSA", "HO-MSSA")) {
  require("rTensor")
  if (identical(kind[1], "HO-SSA")) {
    N <- length(s)
    J <- N - I - L + 2
    X <- outer(1:I, 1:L, `+`) |>
      outer(1:J, function(il, j)
        s[il + j - 2])
  }
  else if (identical(kind[1], "HO-MSSA")) {
    if (missing(L))
      L <- I
    N <- nrow(s)
    K <- N - L + 1
    Q <- ncol(s)
    X <- apply(s, 2, Rssa::hankel, L = L, simplify = FALSE) |>
      Reduce(cbind, x = _) |>
      rTensor::fold(1, 2:3, modes = c(L, K, Q))
  }
  return(as.tensor(X))
}

@@

tucker_mod <- function (tnsr, ranks = NULL, max_iter = 25, tol = 1e-05, status = TRUE) 
{
    stopifnot(is(tnsr, "Tensor"))
    if (is.null(ranks)) 
        stop("ranks must be specified")
    if (sum(ranks > tnsr@modes) != 0) 
        stop("ranks must be smaller than the corresponding mode")
    if (sum(ranks <= 0) != 0) 
        stop("ranks must be positive")
    if (all(tnsr@data == 0)) 
        stop("Zero tensor detected")
    num_modes <- tnsr@num_modes
    U_list <- vector("list", num_modes)
    for (m in 1:num_modes) {
        temp_mat <- rs_unfold(tnsr, m = m)@data
        U_list[[m]] <- svd(temp_mat, nu = ranks[m])$u
    }
    tnsr_norm <- fnorm_complex(tnsr)
    curr_iter <- 1
    converged <- FALSE
    fnorm_resid <- rep(0, max_iter)
    CHECK_CONV <- function(Z, U_list) {
        est <- ttl(Z, U_list, ms = 1:num_modes)
        curr_resid <- fnorm_complex(tnsr - est)
        fnorm_resid[curr_iter] <<- curr_resid
        if (curr_iter == 1) 
            return(FALSE)
        if (abs(curr_resid - fnorm_resid[curr_iter - 1])/tnsr_norm < 
            tol) 
            return(TRUE)
        else {
            return(FALSE)
        }
    }
    if (status) {
      pb <- txtProgressBar(min = 0, max = max_iter, style = 3)
      while ((curr_iter < max_iter) && (!converged)) {
        setTxtProgressBar(pb, curr_iter)
        modes <- tnsr@modes
        modes_seq <- 1:num_modes
        for (m in modes_seq) {
          X <- ttl(tnsr, lapply(U_list[-m], (\(.) Conj(t(.)))), ms = modes_seq[-m])
          U_list[[m]] <- svd(rs_unfold(X, m = m)@data, nu = ranks[m])$u
        }
        Z <- ttm(X, mat = Conj(t(U_list[[num_modes]])), m = num_modes)
        if (CHECK_CONV(Z, U_list)) {
          converged <- TRUE
          setTxtProgressBar(pb, max_iter)
        }
        else {
          curr_iter <- curr_iter + 1
        }
      }
      close(pb)
    }
    else {
      while ((curr_iter < max_iter) && (!converged)) {
        modes <- tnsr@modes
        modes_seq <- 1:num_modes
        for (m in modes_seq) {
          X <- ttl(tnsr, lapply(U_list[-m], (\(.) Conj(t(.)))), ms = modes_seq[-m])
          U_list[[m]] <- svd(rs_unfold(X, m = m)@data, nu = ranks[m])$u
        }
        Z <- ttm(X, mat = Conj(t(U_list[[num_modes]])), m = num_modes)
        if (CHECK_CONV(Z, U_list)) {
          converged <- TRUE
        }
        else {
          curr_iter <- curr_iter + 1
        }
      }
    }

@@

```{r test-hosvd-mod}
library(waldo)
X.test <- rTensor::rand_tensor()
cat("Real example:\n")
waldo::compare(hosvd(X.test), hosvd_mod(X.test))

@@

```{r test-hooi-mod}
library(waldo)
X.test <- rTensor::rand_tensor()
cat("Real example:\n")
waldo::compare(tucker(X.test, c(3, 3, 3)), tucker_mod(X.test, c(3, 3, 3)))

@@

```{r tensor-based-esprit}
tens_esprit <- function(s, I, L, groups, kind = c("HO-SSA", "HO-MSSA"), est_dim = 3, status = TRUE) {
  if (identical(kind[1], "HO-SSA"))
    H <- tens3(s, I, L)
  else if (identical(kind[1], "HO-MSSA"))
  {
    if (missing(L))
      L <- I
    H <- tens3(s, L, kind = "HO-MSSA")
  }
  max_rank <- max(sapply(groups, max))
  H.hooi <- tucker_mod(H, rep(max_rank, 3), status = status)
  estimates <- list()
  for (i in seq(groups)) {
    U <- H.hooi$U[[est_dim]][, groups[[i]], drop = FALSE]
    Z <- qr.solve(U[-nrow(U), ], U[-1, ])
    poles <- eigen(Z, only.values = TRUE)$values
    estimates[[i]] <- list(
      poles = poles,
      rates = Re(log(poles)),
      frequencies = Im(log(poles)) / 2 / pi
    )
  }
  estimates
}
```

@@

```{r rrmse}
rrmse <- function(true, estimate) {
  100 * sqrt(mean(abs(true - estimate)^2)) / true
}
```

@@

>>>>

---
title: "Parameters estimation via HOSVD"
output: html_document
---
```{r setup, include=FALSE}
library(Rssa)
library(rTensor)
library(ggplot2) 
library(dplyr)
library(tidyr)
library(waldo)
```

```{r tensorization}
tens3 <- function(s, I, L, kind = c("HO-SSA", "HO-MSSA")) {
  require("rTensor")
  if (identical(kind[1], "HO-SSA")) {
    N <- length(s)
    J <- N - I - L + 2
    X <- outer(1:I, 1:L, `+`) |>
      outer(1:J, function(il, j)
        s[il + j - 2]) |> as.tensor()
  }
  else if (identical(kind[1], "HO-MSSA")) {
    if (missing(L))
      L <- I
    N <- nrow(s)
    K <- N - L + 1
    Q <- ncol(s)
    X <- apply(s, 2, Rssa::hankel, L = L, simplify = FALSE) |>
      Reduce(cbind, x = _) |>
      rTensor::fold(1, 2:3, modes = c(L, K, Q))
  }
  return(X)
}

@@

tucker_mod <- function(tnsr, ranks = NULL, max_iter = 25, tol = 1e-05, status = TRUE) 
{
    stopifnot(is(tnsr, "Tensor"))
    if (is.null(ranks)) 
        stop("ranks must be specified")
    if (sum(ranks > tnsr@modes) != 0) 
        stop("ranks must be smaller than the corresponding mode")
    if (sum(ranks <= 0) != 0) 
        stop("ranks must be positive")
    if (all(tnsr@data == 0)) 
        stop("Zero tensor detected")
    num_modes <- tnsr@num_modes
    U_list <- vector("list", num_modes)
    for (m in 1:num_modes) {
        temp_mat <- rs_unfold(tnsr, m = m)@data
        U_list[[m]] <- svd(temp_mat, nu = ranks[m])$u
    }
    tnsr_norm <- fnorm_complex(tnsr)
    curr_iter <- 1
    converged <- FALSE
    fnorm_resid <- rep(0, max_iter)
    CHECK_CONV <- function(Z, U_list) {
        est <- ttl(Z, U_list, ms = 1:num_modes)
        curr_resid <- fnorm_complex(tnsr - est)
        fnorm_resid[curr_iter] <<- curr_resid
        if (curr_iter == 1) 
            return(FALSE)
        if (abs(curr_resid - fnorm_resid[curr_iter - 1])/tnsr_norm < 
            tol) 
            return(TRUE)
        else {
            return(FALSE)
        }
    }
    if (status) {
      pb <- txtProgressBar(min = 0, max = max_iter, style = 3)
      while ((curr_iter < max_iter) && (!converged)) {
        setTxtProgressBar(pb, curr_iter)
        modes <- tnsr@modes
        modes_seq <- 1:num_modes
        for (m in modes_seq) {
          X <- ttl(tnsr, lapply(U_list[-m], (\(.) Conj(t(.)))), ms = modes_seq[-m])
          U_list[[m]] <- svd(rs_unfold(X, m = m)@data, nu = ranks[m])$u
        }
        Z <- ttm(X, mat = Conj(t(U_list[[num_modes]])), m = num_modes)
        if (CHECK_CONV(Z, U_list)) {
          converged <- TRUE
          setTxtProgressBar(pb, max_iter)
        }
        else {
          curr_iter <- curr_iter + 1
        }
      }
      close(pb)
    }
    else {
      while ((curr_iter < max_iter) && (!converged)) {
        modes <- tnsr@modes
        modes_seq <- 1:num_modes
        for (m in modes_seq) {
          X <- ttl(tnsr, lapply(U_list[-m], (\(.) Conj(t(.)))), ms = modes_seq[-m])
          U_list[[m]] <- svd(rs_unfold(X, m = m)@data, nu = ranks[m])$u
        }
        Z <- ttm(X, mat = Conj(t(U_list[[num_modes]])), m = num_modes)
        if (CHECK_CONV(Z, U_list)) {
          converged <- TRUE
        }
        else {
          curr_iter <- curr_iter + 1
        }
      }
    }

@@

```{r test-hosvd-mod}
X.test <- rTensor::rand_tensor()
cat("Real example:\n")
waldo::compare(hosvd(X.test), hosvd_mod(X.test))

@@

```{r test-hooi-mod}
X.test <- rTensor::rand_tensor()
cat("Real example:\n")
waldo::compare(tucker(X.test, c(3, 3, 3)), tucker_mod(X.test, c(3, 3, 3)))

@@

```{r tensor-based-esprit}
tens_esprit <- function(s, I, L, groups, kind = c("HO-SSA", "HO-MSSA"), est_dim = 3, 
                        status = TRUE, qrtol = 1e-07) {
  if (identical(kind[1], "HO-SSA"))
    H <- tens3(s, I, L)
  else if (identical(kind[1], "HO-MSSA"))
  {
    if (missing(L))
      L <- I
    H <- tens3(s, L, kind = "HO-MSSA")
  }
  max_rank <- max(sapply(groups, max))
  H.hooi <- tucker_mod(H, rep(max_rank, 3), status = status)
  estimates <- list()
  for (i in seq(groups)) {
    U <- H.hooi$U[[est_dim]][, groups[[i]], drop = FALSE]
    Z <- qr.solve(U[-nrow(U), ], U[-1, ], qrtol)
    poles <- eigen(Z, only.values = TRUE)$values
    estimates[[i]] <- list(
      poles = poles,
      rates = Re(log(poles)),
      frequencies = Im(log(poles)) / 2 / pi
    )
  }
  estimates
}
```

@@

```{r rrmse}
rrmse <- function(true, estimate) {
  100 * sqrt(mean(abs(true - estimate)^2)) / abs(true)
}
```

@@

```{r 1d-comparison-for-sds, results='hide'}
len.out <- 16
sds <- seq(from = 0.01, to = 0.06, length.out = len.out)
set.seed(10)

n <- 25
L <- 15
I <- 14; L.tens <- 8
rates.true <- c(-0.01, -0.02)
freqs.true <- c(0.2, 0.22)
signal <- exp((rates.true[1] + 2i * pi * freqs.true[1]) * 0:(n - 1)) +
  exp((rates.true[2] + 2i * pi * freqs.true[2]) * 0:(n - 1))
R <- 1000

rate1.rrmse <- list(ssa = numeric(len.out), tssa = numeric(len.out))
rate2.rrmse <- list(ssa = numeric(len.out), tssa = numeric(len.out))
freq1.rrmse <- list(ssa = numeric(len.out), tssa = numeric(len.out))
freq2.rrmse <- list(ssa = numeric(len.out), tssa = numeric(len.out))

pb <- txtProgressBar(max = R * len.out, style = 3)

for (j in seq_along(sds)) {
  pars.ssa <- list(
    rate1 = numeric(R),
    rate2 = numeric(R),
    freq1 = numeric(R),
    freq2 = numeric(R)
  )

  pars.tssa <- list(
    rate1 = numeric(R),
    rate2 = numeric(R),
    freq1 = numeric(R),
    freq2 = numeric(R)
  )

  sd <- sds[j]

  for (i in seq(R)) {
    noise <- CCSWGN(n, mean = 0, sd = sd)
    x <- signal + noise
    x.ssa <- ssa(x, L = L)

    par.ssa <- parestimate(x.ssa, groups = list(1:2))
    pars.ssa$rate1[i] <- max(par.ssa$rates[1], par.ssa$rates[2])
    pars.ssa$rate2[i] <- min(par.ssa$rates[1], par.ssa$rates[2])
    pars.ssa$freq1[i] <- min(par.ssa$frequencies[1], par.ssa$frequencies[2])
    pars.ssa$freq2[i] <- max(par.ssa$frequencies[1], par.ssa$frequencies[2])

    par.tssa <- tens_esprit(x, I, L.tens, groups = list(1:2), status = FALSE)
    pars.tssa$rate1[i] <- max(par.tssa[[1]]$rates[1], par.tssa[[1]]$rates[2])
    pars.tssa$rate2[i] <- min(par.tssa[[1]]$rates[1], par.tssa[[1]]$rates[2])
    pars.tssa$freq1[i] <- min(par.tssa[[1]]$frequencies[1], par.tssa[[1]]$frequencies[2])
    pars.tssa$freq2[i] <- max(par.tssa[[1]]$frequencies[1], par.tssa[[1]]$frequencies[2])

    setTxtProgressBar(pb, (j - 1) * R + i)
  }

  rate1.rrmse$ssa[j] <- rrmse(rates.true[1], pars.ssa$rate1)
  rate1.rrmse$tssa[j] <- rrmse(rates.true[1], pars.tssa$rate1)
  rate2.rrmse$ssa[j] <- rrmse(rates.true[2], pars.ssa$rate2)
  rate2.rrmse$tssa[j] <- rrmse(rates.true[2], pars.tssa$rate2)
  freq1.rrmse$ssa[j] <- rrmse(freqs.true[1], pars.ssa$freq1)
  freq1.rrmse$tssa[j] <- rrmse(freqs.true[1], pars.tssa$freq1)
  freq2.rrmse$ssa[j] <- rrmse(freqs.true[2], pars.ssa$freq2)
  freq2.rrmse$tssa[j] <- rrmse(freqs.true[2], pars.tssa$freq2)
}
close(pb)
```

```{r plot-and-save}
plot.save <- function(plt, name, format, ...) {
  print(plt)
  format(name, ...)
  print(plt)
  dev.off()
}
```

```{r}
rate1.df <- rate1.rrmse |> as.data.frame() |> mutate(sd = sds) |>
  pivot_longer(!sd, names_to = "type", values_to = "RRMSE")
rate1.plot <- rate1.df |> ggplot() +
  geom_line(aes(
    x = sd,
    y = RRMSE,
    colour = type,
    linetype = type
  ), lwd = 1.2)

plot.save(rate1.plot,
          "./img/rate1_mini.pdf",
          pdf,
          width = 9,
          height = 6)
```

```{r}
rate2.df <- rate2.rrmse |> as.data.frame() |> mutate(sd = sds) |>
  pivot_longer(!sd, names_to = "type", values_to = "RRMSE")
rate2.plot <- rate2.df |> ggplot() + geom_line(aes(
  x = sd,
  y = RRMSE,
  colour = type,
  linetype = type
), lwd = 1.2)

plot.save(rate2.plot,
          "./img/rate2_mini.pdf",
          pdf,
          width = 9,
          height = 6)
```

```{r}
freq1.df <- freq1.rrmse |> as.data.frame() |> mutate(sd = sds) |>
  pivot_longer(!sd, names_to = "type", values_to = "RRMSE")
freq1.plot <- freq1.df |> ggplot() + geom_line(aes(
  x = sd,
  y = RRMSE,
  colour = type,
  linetype = type
), lwd = 1.2)

plot.save(freq1.plot,
          "./img/freq1_mini.pdf",
          pdf,
          width = 9,
          height = 6)
```
```{r}
freq2.df <- freq2.rrmse |> as.data.frame() |> mutate(sd = sds) |>
  pivot_longer(!sd, names_to = "type", values_to = "RRMSE")
freq2.plot <- freq2.df |> ggplot() + geom_line(aes(
  x = sd,
  y = RRMSE,
  colour = type,
  linetype = type
), lwd = 1.2)

plot.save(freq2.plot,
          "./img/freq2_mini.pdf",
          pdf,
          width = 9,
          height = 6)
```

```{r complex-nd-esprit}
cmesprit <- function(s, L, groups, qrtol = 1e-07) {
  H <- apply(s, 2, Rssa::hankel, L = L, simplify = FALSE) |>
    Reduce(cbind, x = _)

  max_rank <- max(sapply(groups, max))
  H.svd <- svd(H, nu = max_rank, nv = 0)
  estimates <- list()
  for (i in seq(groups)) {
    U <- H.svd$u[, groups[[i]], drop = FALSE]
    Z <- qr.solve(U[-nrow(U), ], U[-1, ], qrtol)
    poles <- eigen(Z, only.values = TRUE)$values
    estimates[[i]] <- list(
      poles = poles,
      rates = Re(log(poles)),
      frequencies = Im(log(poles)) / 2 / pi
    )
  }
  estimates
}
```

```{r test-nd-ssa-esprit-no-noise}
set.seed(10)

n <- 25
Q <- 12
coef <- matrix(CCSWGN(2 * Q), ncol = 2)

x <- exp((-0.01 + 2i * pi * 0.2) * 0:(n - 1)) %o% coef[, 1] + 
  exp((-0.02 + 2i * pi * 0.22) * 0:(n - 1)) %o% coef[, 2]
L <- 13
pars <- cmesprit(x, L = L, groups = list(1:2))[[1]]
cat("Rates:\n", pars$rates, "\nFrequencies:\n", pars$frequencies)
```

```{r test-nd-tens-esprit-no-noise}
set.seed(10)

n <- 25
Q <- 12
coef <- matrix(CCSWGN(2 * Q), ncol = 2)

x <- exp((-0.01 + 2i * pi * 0.2) * 0:(n - 1)) %o% coef[, 1] + 
  exp((-0.02 + 2i * pi * 0.22) * 0:(n - 1)) %o% coef[, 2]
L <- 13
pars <- tens_esprit(x, I, L, groups = list(1:2), kind = "HO-MSSA", est_dim = 1, status = FALSE)
cat("Rates:\n", pars[[1]]$rates, "\nFrequencies:\n", pars[[1]]$frequencies)
```

```{r nd-comparison-for-sds}
len.out <- 16
sds <- seq(from = 0.1, to = 0.4, length.out = len.out)
set.seed(10)

n <- 25
Q <- 12
L <- 13
L.tens <- 13

rates.true <- c(-0.01, -0.02)
freqs.true <- c(0.2, 0.22)
coef <- matrix(CCSWGN(2 * Q), ncol = 2)

signal <- exp((rates.true[1] + 2i * pi * freqs.true[1]) * 0:(n - 1)) %o% coef[, 1] + 
  exp((rates.true[2] + 2i * pi * freqs.true[2]) * 0:(n - 1)) %o% coef[, 2]
R <- 1000

rate1.rrmse <- list(ssa = numeric(len.out), tssa = numeric(len.out))
rate2.rrmse <- list(ssa = numeric(len.out), tssa = numeric(len.out))
freq1.rrmse <- list(ssa = numeric(len.out), tssa = numeric(len.out))
freq2.rrmse <- list(ssa = numeric(len.out), tssa = numeric(len.out))

pb <- txtProgressBar(max = R * len.out, style = 3)

for (j in seq_along(sds)) {
  pars.ssa <- list(
    rate1 = numeric(R),
    rate2 = numeric(R),
    freq1 = numeric(R),
    freq2 = numeric(R)
  )

  pars.tssa <- list(
    rate1 = numeric(R),
    rate2 = numeric(R),
    freq1 = numeric(R),
    freq2 = numeric(R)
  )

  sd <- sds[j]

  for (i in seq(R)) {
    noise <- CCSWGN(n, mean = 0, sd = sd)
    x <- signal + noise

    par.ssa <- cmesprit(x, L = L, groups = list(1:2))
    pars.ssa$rate1[i] <- max(par.ssa[[1]]$rates[1], par.ssa[[1]]$rates[2])
    pars.ssa$rate2[i] <- min(par.ssa[[1]]$rates[1], par.ssa[[1]]$rates[2])
    pars.ssa$freq1[i] <- min(par.ssa[[1]]$frequencies[1], par.ssa[[1]]$frequencies[2])
    pars.ssa$freq2[i] <- max(par.ssa[[1]]$frequencies[1], par.ssa[[1]]$frequencies[2])

    par.tssa <- tens_esprit(x, L.tens, kind = "HO-MSSA", est_dim = 1, groups = list(1:2), status = FALSE)
    pars.tssa$rate1[i] <- max(par.tssa[[1]]$rates[1], par.tssa[[1]]$rates[2])
    pars.tssa$rate2[i] <- min(par.tssa[[1]]$rates[1], par.tssa[[1]]$rates[2])
    pars.tssa$freq1[i] <- min(par.tssa[[1]]$frequencies[1], par.tssa[[1]]$frequencies[2])
    pars.tssa$freq2[i] <- max(par.tssa[[1]]$frequencies[1], par.tssa[[1]]$frequencies[2])

    setTxtProgressBar(pb, (j - 1) * R + i)
  }

  rate1.rrmse$ssa[j] <- rrmse(rates.true[1], pars.ssa$rate1)
  rate1.rrmse$tssa[j] <- rrmse(rates.true[1], pars.tssa$rate1)
  rate2.rrmse$ssa[j] <- rrmse(rates.true[2], pars.ssa$rate2)
  rate2.rrmse$tssa[j] <- rrmse(rates.true[2], pars.tssa$rate2)
  freq1.rrmse$ssa[j] <- rrmse(freqs.true[1], pars.ssa$freq1)
  freq1.rrmse$tssa[j] <- rrmse(freqs.true[1], pars.tssa$freq1)
  freq2.rrmse$ssa[j] <- rrmse(freqs.true[2], pars.ssa$freq2)
  freq2.rrmse$tssa[j] <- rrmse(freqs.true[2], pars.tssa$freq2)
}
close(pb)
```

```{r}
rate1.df <- rate1.rrmse |> as.data.frame() |> mutate(sd = sds) |>
  pivot_longer(!sd, names_to = "type", values_to = "RRMSE")
rate1.plot <- rate1.df |> ggplot() +
  geom_line(aes(
    x = sd,
    y = RRMSE,
    colour = type,
    linetype = type
  ), lwd = 1.2)

plot.save(rate1.plot,
          "./img/rate1_nd.pdf",
          pdf,
          width = 9,
          height = 6)
```

```{r}
rate2.df <- rate2.rrmse |> as.data.frame() |> mutate(sd = sds) |>
  pivot_longer(!sd, names_to = "type", values_to = "RRMSE")
rate2.plot <- rate2.df |> ggplot() + geom_line(aes(
  x = sd,
  y = RRMSE,
  colour = type,
  linetype = type
), lwd = 1.2)

plot.save(rate2.plot,
          "./img/rate2_nd.pdf",
          pdf,
          width = 9,
          height = 6)
```

```{r}
freq1.df <- freq1.rrmse |> as.data.frame() |> mutate(sd = sds) |>
  pivot_longer(!sd, names_to = "type", values_to = "RRMSE")
freq1.plot <- freq1.df |> ggplot() + geom_line(aes(
  x = sd,
  y = RRMSE,
  colour = type,
  linetype = type
), lwd = 1.2)

plot.save(freq1.plot,
          "./img/freq1_nd.pdf",
          pdf,
          width = 9,
          height = 6)
```
```{r}
freq2.df <- freq2.rrmse |> as.data.frame() |> mutate(sd = sds) |>
  pivot_longer(!sd, names_to = "type", values_to = "RRMSE")
freq2.plot <- freq2.df |> ggplot() + geom_line(aes(
  x = sd,
  y = RRMSE,
  colour = type,
  linetype = type
), lwd = 1.2)

plot.save(freq2.plot,
          "./img/freq2_nd.pdf",
          pdf,
          width = 9,
          height = 6)
```
